Finally Tagless, Partially Evaluated

### 2008-09-03 tagless/tagless-final.tex
some more work on JFP paper (some real advance but still too many notes)
A few comments and changes

\begin{enumerate}
\item We build interpreters that evaluate (\S\ref{language}),
   compile (or evaluate with staging) (\S\ref{S:compiler}), and 
   partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing
   styles\ifshort\else\ (\S\ref{variations})\fi.
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\ifshort\else
\item We show the first typed, tagless, and statically type-preserving CPS
    transformation with simple types (\S\ref{variations}), which is available for mainstream
    functional languages (ML and Haskell). Type preservation is assured by
    the soundness of our embedding.
\fi
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our
  interpreters, from evaluation and compilation (\S\ref{language}) 
   to partial evaluation \ifshort\else and CPS transformation \fi(\S\ref{PE}).
\item We point a clear way to extend the object language with more features
    such as state\ifshort\else~(\S\ref{state})\fi.\ifshort\footnote{Again, please see our code.}\fi
    \ Our term encoding is contravariant in the object language, so
    extending the language does not invalidate terms already encoded.
\item We describe an approach to self\hyp interpretation compatible with the
  above\ifshort\else~(\S\ref{selfinterp})\fi.  Self\hyp interpretation turned
  out to be harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
\end{enumerate}
Our code is surprisingly simple and obvious in hindsight, but
it has been cited as a difficult problem (\cite{sumii-hybrid}
notwithstanding) to interpret a typed object language in a typed metalanguage
without tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

@@

@@

Figure~\ref{fig:object} shows our object language, a simply-typed
$\lambda$-calculus with fixpoint, integers, booleans, and comparison.
The language is close to \citets{xi-guarded}, without their polymorphic
lift but with more constants so as to more conveniently express Fibonacci,
factorial, and power.
\begin{comment}
\oleg{There is now IncopeDB.hs that defines at least R, L and C
 interpreters (no PE at present though) using de Bruijn indices. So we
 can say that (some of) our development is available in de Bruijn
 style. We won't talk about it in the paper though.}
\end{comment}
In contrast to \S\ref{intro}, we use higher-order
abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices to encode binding and
ensures that our object programs are closed.
We find HOAS to be more convenient, though we could have continued to
use de Bruijn indices.

@@

  let int  (x:int)  = x
  let bool (b:bool) = b
  let lam  f        = f
  let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self
  let add  e1 e2    = e1 + e2
  let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then et () else ee ()
end
\end{code}
\fi
%
As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding}
with~|R|\ifshort: \texttt{module EXR = EX(R)}\fi.
\ifshort\else
\begin{code}
module EXR = EX(R)
\end{code}
\fi
Thus, |EXR.test1 ()| evaluates to the untagged boolean value |true|.
It
\begin{comment}
%% commenting out this Haskell in JFP version as well
\ifshort It \else
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
In OCaml, too, it
\fi 
\enc{comment}
is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
\ifshort
(The source code shows a total interpreter |L| that measures
the size of each object term.)
\fi
\ifshort
We can also generalize from~|R| to all interpreters;
these propositions follow immediately from the soundness of the
metalanguage's type system.
\fi
\begin{proposition}
If an object term~$e$ encoded (as $e'$) in the metalanguage  has (object)
type~$t$, then evaluating~$e'$ in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~$t$.
\end{proposition}
\ifshort\else
Generalizing from~|R| to all interpreters, we have the following
broader and more useful, if also more obvious, proposition.
\fi
\begin{proposition}
  If, for an implementation |I| of |Symantics|, if for all object
  terms $e$ (encoded as $e'$), evaluating $e'$ with |I| always
  terminates, then 
  the type system of the object
  language is sound with respect to the dynamic semantics defined by
  that implementation.
  In other words, the (total) relation induced by |I| is a sub-relation
  of the one expressed by the type system of the object language.
  \jacques{Please check that I have this right}
\end{proposition}
\ifshort\else
These propositions follow immediately from the soundness of the
metalanguage's type system.
\fi
\ifshort\else
For variety, we show another interpreter, which measures the \emph{size}
(or \emph{length}, thus the mnemonic |L|)
of each object term, defined as the number of term
constructors.
\begin{code}
module L = struct
  type ('c,'dv) repr = int

@@

To enable these simplifications, we delay binding-time analysis
for a static function until it is applied, that is, until |lam f|
appears as the argument of |app|.  To do so, we have to incorporate |f|
as it is into the |P1.repr| data structure: the representation
for a function type |'a->'b| should be one of
\begin{code}
S1 of ('c,'a) repr -> ('c,'b) repr | E1 of ('c,'a->'b) C.repr
P1 of (('c,'a) repr -> ('c,'b) repr) option * ('c,'a->'b) C.repr
\end{code}
unlike |P1.repr| of |int| or |bool|.
That is, we need a nonparametric data type, something akin to
type-indexed functions and type-indexed types, which
\citet{oliveira-typecase} dub the \emph{typecase} design pattern.
Thus, typed partial evaluation, like typed CPS transformation
(see \S\ref{S:CPS}),
inductively defines a map from source types to target types that
performs case distinction on the source type. In Haskell, typecase
can be implemented using either GADTs or
type-class functional dependencies
\citep{oliveira-typecase}. The accompanying code shows both
approaches (|Incope.hs| and |incope1.hs|), 
neither of which is portable to OCaml. In addition,
the problem of non\hyp exhaustive pattern\hyp matching reappears in
the GADT approach because GHC 6.6.1 cannot see that a particular
type of a GADT value precludes certain constructors.
Thus GADTs fail to
make it \emph{syntactically} apparent that pattern matching is exhaustive.
\jacques{what about GHC 6.8?}

@@

Figure~\ref{fig:pe} shows the partial evaluator~|P|.
Its type |repr| expresses the type definition for |repr_pe| given
as the start of this section, with |'sv| taking the place of |repr_pe t|.
The function |abstr|
extracts a future-stage code value from the result of
partial evaluation.  Conversely, the function |pdyn| injects a
code value into the |repr| type. As
in~\S\ref{S:PE-problem}, we build dynamic terms alongside
any static ones to avoid polymorphic lift.

@@

Given an \emph{encoding} of each object term~$e$ as an \emph{object} term~$\Encode{e}$,
a \emph{self\hyp interpreter} is usually defined as an object
function~$\si$ such that any object term~$e$ is observationally
equivalent to the object application $\si\Encode{e}$
\citep{jones-partial,taha-tag,Danvy-tagging-encoding}.
A partial evaluator~$\pe$ maps object terms~$e$ to observationally
equivalent object terms~$\pe(e)$.  It is said to be
\emph{optimal} with respect to~|si| \citep{jones-challenging} if $\pe(\si\Encode{e})$
is $\alpha$\hyp equivalent to~$e$ (or in some accounts, no
less efficient than~$e$).

>>>>

\begin{enumerate}
\item We build interpreters that evaluate (\S\ref{language}),
   compile (or evaluate with staging) (\S\ref{S:compiler}), and 
   partially evaluate (\S\ref{PE}) a typed higher-order object language
   in a typed metalanguage, in direct and continuation\hyp passing
   styles\ifshort\else\ (\S\ref{variations})\fi.
\item All these interpreters use no type tags, patently never get stuck,
    and need no advanced type-system features such as GADTs, dependent types,
    or intentional type analysis.
\item The partial evaluator avoids polymorphic lift and delays binding-time
    analysis.  It bakes a type-to-type map into the interpreter
    interface to eliminate the need for GADTs and thus remain portable
    across Haskell 98 and ML.
\ifshort\else
\item We show the first typed, tagless, and statically type-preserving CPS
    transformation with simple types (\S\ref{variations}), which is available for mainstream
    functional languages (ML and Haskell). Type preservation is assured by
    the soundness of our embedding.
\fi
\item We use the type system of the metalanguage
    to check statically that an object program is well-typed and closed.
\item We show clean, comparable implementations in MetaOCaml and Haskell.
\item We specify a functor signature that encompasses all our
  interpreters, from evaluation and compilation (\S\ref{language}) 
   to partial evaluation \ifshort\else and CPS transformation \fi(\S\ref{PE}).
\item We point a clear way to extend the object language with more features
    such as state\ifshort\else~(\S\ref{state})\fi.\ifshort\footnote{Again, please see our code.}\fi
    \ Our term encoding is contravariant in the object language, so
    extending the language does not invalidate terms already encoded.
\item We describe an approach to self\hyp interpretation compatible with the
  above\ifshort\else~(\S\ref{selfinterp})\fi.  Self\hyp interpretation turned
  out to be harder than expected.\ifshort\footnotemark[\value{footnote}]\fi
\item We show how the module system of Ocaml and Haskell's type class 
   system can be used to parametrize the semantics of (a particular style of
   syntax encoding of) lamda terms.
\end{enumerate}
Our code is surprisingly simple and obvious in hindsight, but
it has been cited as a difficult problem (\cite{sumii-hybrid}
notwithstanding) to interpret a typed object language in a typed metalanguage
without tagging or type\hyp system extensions.  For example, \citet{taha-tag}
say that ``expressing such an interpreter in a statically typed
programming language is a rather subtle matter. In fact, it is only
recently that some work on programming type-indexed values in ML
\citep{yang-encoding} has given a hint of how such a function can be
expressed.''  We discuss related work in~\S\ref{related}.

@@

A referee aptly summarized our paper as ``proposing a way to write a 
typed fold function over a typed term''.

@@

Figure~\ref{fig:object} shows our object language, a simply-typed
$\lambda$-calculus with fixpoint, integers, booleans, and comparison.
The language is close to \citets{xi-guarded}, without their polymorphic
lift but with more constants so as to more conveniently express Fibonacci,
factorial, and power.
In contrast to \S\ref{intro}, in the rest of the paper 
we use higher-order
abstract syntax (HOAS) \citep{miller-manipulating,pfenning-higher-order}
rather than de Bruijn indices to encode binding and
ensures that our object programs are closed.
We find HOAS to be more convenient. We do present the version of our
developement using de Bruijn indices in the accompanying source code, file
|incope-dB.ml|.

@@

  let int  (x:int)  = x
  let bool (b:bool) = b
  let lam  f        = f
  let app  e1 e2    = e1 e2
  let fix  f        = let rec self n = f self n in self
  let add  e1 e2    = e1 + e2
  let mul  e1 e2    = e1 * e2
  let leq  x y      = x <= y
  let if_  eb et ee = if eb then et () else ee ()
end
\end{code}
\fi
%
As in~\S\ref{ourapproach},
this interpreter is patently tagless, using neither a universal type nor
any pattern matching: the operation |add| is really
OCaml's addition, and |app| is OCaml's application. To run our
examples, we instantiate the |EX| functor from~\S\ref{encoding}
with~|R|\ifshort: \texttt{module EXR = EX(R)}\fi.
\ifshort\else
\begin{code}
module EXR = EX(R)
\end{code}
\fi
Thus, |EXR.test1 ()| evaluates to the untagged boolean value |true|.
\begin{comment}
%% commenting out this Haskell in JFP version as well
In Haskell, we define
\begin{code}
newtype R a = R {unR::a}
instance Symantics R where ...
\end{code}
Although |R| looks like a tag, it is only
a |newtype|.  The types |a| and |R a| are represented differently
only at compile time, not at run time.  Pattern matching against~|R|
cannot ever fail and is assuredly compiled away.
\end{comment}
It is obvious to the compiler that
pattern matching cannot fail, because there is no
pattern matching. Evaluation can only fail to yield a value
due to interpreting |fix|.
\ifshort
(The source code shows a total interpreter |L| that measures
the size of each object term.)
\fi
\ifshort
We can also generalize from~|R| to all interpreters;
these propositions follow immediately from the soundness of the
metalanguage's type system.
\fi
\begin{proposition}
If an object term~$e$ encoded (as $e'$) in the metalanguage  has (object)
type~$t$, then evaluating~$e'$ in the interpreter~|R| either continues
indefinitely or terminates with a value of the same type~$t$.
\end{proposition}
\ifshort\else
Generalizing from~|R| to all interpreters, we have the following
broader and more useful, if also more obvious, proposition.
\fi
\begin{proposition}
  If, for an implementation |I| of |Symantics|, if for all object
  terms $e$ (encoded as $e'$), evaluating $e'$ with |I| always
  terminates, then 
  the type system of the object
  language is sound with respect to the dynamic semantics defined by
  that implementation.
  In other words, the (total) relation induced by |I| is a sub-relation
  of the one expressed by the type system of the object language.
  \jacques{Please check that I have this right}
\end{proposition}
\ifshort\else
These propositions follow immediately from the soundness of the
metalanguage's type system.
\fi
\ifshort\else
For variety, we show another interpreter, which measures the \emph{size}
(or \emph{length}, thus the mnemonic |L|)
of each object term, defined as the number of term
constructors.
\begin{code}
module L = struct
  type ('c,'dv) repr = int

@@

To enable these simplifications, we delay binding-time analysis
for a static function until it is applied, that is, until |lam f|
appears as the argument of |app|.  To do so, we have to incorporate |f|
as it is into the |P1.repr| data structure: the representation
for a function type |'a->'b| should be one of
\begin{code}
S1 of ('c,'a) repr -> ('c,'b) repr | E1 of ('c,'a->'b) C.repr
P1 of (('c,'a) repr -> ('c,'b) repr) option * ('c,'a->'b) C.repr
\end{code}
unlike |P1.repr| of |int| or |bool|.
That is, we need a nonparametric data type, something akin to
type-indexed functions and type-indexed types, which
\citet{oliveira-typecase} dub the \emph{typecase} design pattern.
Thus, typed partial evaluation, like typed CPS transformation
(see \S\ref{S:CPS}),
inductively defines a map from source types to target types that
performs case distinction on the source type. In Haskell, typecase
can be implemented using either GADTs or
type-class functional dependencies
\citep{oliveira-typecase}. The accompanying code shows both
approaches (|Incope.hs| and |incope1.hs|), 
neither of which is portable to OCaml. In addition,
the problem of non\hyp exhaustive pattern\hyp matching reappears in
the GADT approach because GHC 6.8 and prior cannot see that a particular
type of a GADT value precludes certain constructors. Although this
is the implementation problem of GHC,
it indicates that assuring exhaustive pattern match with GADTs
requires non-trivial reasoning (beyond the abilities of GHC at the moment);
thus GADTs fail to
make it \emph{syntactically} apparent that pattern matching is exhaustive.

@@

Figure~\ref{fig:pe} shows the partial evaluator~|P|.
Its type |repr| expresses the type definition for |repr_pe| given
at the start of this section, with |'sv| taking the place of |repr_pe t|.
The function |abstr|
extracts a future-stage code value from the result of
partial evaluation.  Conversely, the function |pdyn| injects a
code value into the |repr| type. As
in~\S\ref{S:PE-problem}, we build dynamic terms alongside
any static ones to avoid polymorphic lift.

@@

Given an \emph{encoding} of each object term~$e$ as an \emph{object}
term~$\Encode{e}$, a \emph{self\hyp interpreter} is usually defined as
an object function~$\si$ such that any object term~$e$ is
observationally equivalent to the object application $\si\Encode{e}$
\citep{jones-partial,taha-tag,Danvy-tagging-encoding}.  A particular
use of self-interpreters is in defining the notion of optimality of
partial evaluation.  A partial evaluator~$\pe$ maps object terms~$e$
to observationally equivalent object terms~$\pe(e)$.  Recall that
$\si\Encode{e}$ is too observationally equivalent to $e$ but less
efficient (``observing'' $\si\Encode{e}$ takes more resources). A
partial evaluator is said to be \emph{optimal} with respect to~|si|
\citep{jones-challenging} if the partial evaluator removes all the 
ineffiency in $\si\Encode{e}$ added by the self-interpreter: that is,
for all $e$, $\pe(\si\Encode{e})$ is $\alpha$\hyp
equivalent to~$e$ (or in some accounts, no less efficient than~$e$).
